A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching

### 2016-10-25 JTB_ms/JTB_ms.Rmd
finish drafting main text

3. _Provide well-documented, open-access code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide open access to the model in the form of detailed, commented code written in the R language [@RALanguageandEn:2014wf]. With an accessible and modular framework, we envision this as a resource for futher development by the scientific community to include additional complexity and problem-specific components. To widen the audience for this work, we chose the R language because it is freely available and in common use by biologists and ecologists.

@@

# Model description
In this dynamic bioenergetic model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1, and each model flux and parameter are defined in Tables 1 and 2, respectively. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Host biomass ($H$), symbiont biomass ($S$), and prey biomass ($X$) have fixed, but different, molar N:C ratios (Table 2). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the parallel complementary formulation of @Kooijman:2010vd to specify these fluxes. The two state variables are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area, and surface area is proportional to volume (i.e., corals are "V1-morphs" in DEB terminology [@Kooijman:2010vd]), biomass increases exponentially during growth (indeed, corals grow exponentially [@Bak:1976bv]). The specific growth rate and the ratio of symbiont to host biomass are the responses of interest of the system. Below we describe the formulation of each flux involved in producing these responses.

@@

| Symbol    | Description                            | Units                | Eq. no. |
| --------- | -------------------------------------- | -------------------- | ------- |
| $j_X$     | Prey uptake rate                       | molX CmolH^-1^ d^-1^ | 3 |
| $j_N$     | Nitrogen uptake rate                   | molN CmolH^-1^ d^-1^ | 4 |
| $j_{HG}$  | Host biomass formation rate            | CmolH CmolH^-1^ d^-1^| 5 |
| $r_{NH}$  | Recycled nitrogen from host turnover   | molN CmolH^-1^ d^-1^ | 6 |
| $\rho_N$  | Nitrogen shared with the symbiont      | molN CmolH^-1^ d^-1^ | 7 |
| $j_{eC}$  | Excess carbon used to activate host CCMs| molC CmolH^-1^ d^-1^ |  |
| $j_L$     | Light absorption rate                  | molph CmolS^-1^ d^-1^| 9 |
| $j_{CO_2}$| CO~2~ input to photosynthesis          | molCO~2~ CmolH^-1^ d^-1^| 10 |
| $r_{CH}$  | Recycled CO~2~ from host turnover      | molCO~2~ CmolH^-1^ d^-1^| 11 |
| $r_{CS}$  | Recycled CO~2~ from symbiont turnover  | molCO~2~ CmolS^-1^ d^-1^| 12 |
| $j_{CP}$  | Photosynthesis rate                    | molC CmolS^-1^ d^-1^ | 13 |
| $j_{eL}$  | Light energy in excess of photochemistry| molph CmolS^-1^ d^-1^| 14 |
| $j_{NPQ}$ | Total capacity of NPQ                  | molph CmolS^-1^ d^-1^ | 15 |
| $c_{ROS}$ | ROS production proportional to baseline | -- | 16 |
| $r_{NS}$  | Recycled nitrogen from symbiont turnover| molN CmolS^-1^ d^-1^| 17 |
| $j_{SG}$  | Symbiont biomass formation rate        | CmolS CmolS^-1^ d^-1^| 18 |
| $\rho_C$  | Fixed carbon shared with host          | molC CmolS^-1^ d^-1^ | 19 |
| $j_{ST}$  | Symbiont turnover rate                 | CmolS CmolS^-1^ d^-1^| 20 |

@@

| Symbol  | Description                            | Value  | Units              |
| ------- | -------------------------------------- | ------ | ------------------ |
| $n_{NH}$ | N:C molar ratio in host biomass | 0.19 | -- |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | 0.2 | -- |
| $n_{NX}$ | N:C molar ratio in prey biomass | 0.13 | -- |
| $j_{HT}^0$ | Specific turnover rate of host biomass | 0.03 | CmolH CmolH^-1^ d^-1^ |
| $j_{ST}^0$ | Specific turnover rate of symbiont biomass | 0.03 | CmolS CmolS^-1^ d^-1^ |
| $\sigma_{NH}$ | Proportion host nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CH}$ | Proportion host carbon turnover recycled | 0.9 | -- |
| $\sigma_{NS}$ | Proportion symbiont nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CS}$ | Proportion symbiont carbon turnover recycled | 0.9 | -- |
| $j_{Xm}$ | Maximum specific feeding rate of host | 0.1292 | molX CmolH^-1^ d^-1^ |
| $K_X$ | Half-saturation constant for prey uptake by host | 20e-6 | molX L^-1^ | 
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | 0.048 | molN CmolH^-1^ d^-1^ |
| $K_N$ | Half-saturation constant for DIN uptake by host | 0.46e-6 | molN L^-1^ | 
| $k_{CO_2}$ | Efficacy of CO~2~ delivery by host CCMs | 10 | molCO~2~ molC^-1^ |
| $j_{HGm}$ | Maximum specific growth rate of host | 1 | CmolH CmolH^-1^ d^-1^ |
| $y_{CL}$ | Quantum yield of photosynthesis | 0.1 | molC mol ph^-1^ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | 1.34 | m^2^ CmolS^-1^ |
| $k_{NPQ}$ | NPQ capacity of symbiont | 112 | mol ph CmolS^-1^ d^-1^ |
| $k_{ROS}$ | Excess photon energy that doubles ROS production, relative to baseline levels | 40-80 | mol ph CmolS^-1^ d^-1^ |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | 2.8 | molC CmolS^-1^ d^-1^ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | 0.25 | CmolS CmolS^-1^ d^-1^ |
| $b$ | Scaling parameter for strength of bleaching response | 5 | -- |

@@

We then specify the input of CO~2~ to photosynthesis as $j_{CO_2}$, a flux mediated by the host that encompasses potentially diverse carbon-concentrating mechanisms (CCMs), including active transport of bicarbonate, carbonic anhydrase-catalyzed conversion of bicarbonate to CO~2~ to promote diffusion toward the symbiont [@Tansik:2015cm], and acidification of the symbiosome to increase localized CO~2~ concentrations around the symbiont [@Barott:2014gx]. Since these mechanisms require energetic input by the host, the magnitude of CO~2~ delivery is proportional to the host's excess carbon, $j_{eC}$, assuming that some of this carbon is respired to energize these CCMs. This formulation means that the symbiont indirectly ensures its own CO~2~ supply by providing fixed carbon to the host [@Wooldridge:2013tj], which establishes a positive feedback in the model that leads to the existence of multiple stable states, which we discuss below in the context of coral bleaching. The parameter $k_{CO_2}$ scales the efficacy of host CCMs, which enables the comparison of different rates of CO~2~ delivery that may characterize different coral species [@Wooldridge:2014di]. The input of CO~2~ to the photosynthesis SU is therefore specified as:

@@

# Results
## Qualitative dynamics
In a constant environment, the system ultimately reaches a steady state of exponential growth or decline. Under some conditions, alternate steady states can be reached depending on initial values of symbiont and host biomass. The mechanism that produces these alternate steady states is the positive feedback between carbon-limitation of the host and CO~2~-limitation of photosynthesis. If the host becomes carbon-limited, the efficacy of its CCMs are reduced and photosynthesis becomes progressively more CO~2~-limited. In turn, less photosynthate is translocated to the host, causing further carbon-limitation and an ultimate decline in biomass; this steady state represents a dysfunctional symbiosis. However, if the host is not carbon-limited, it effectively delivers CO~2~ to photosynthesis and ensures its continued receipt of fixed carbon from the symbiont; this system is nitrogen-limited and represents a functional symbiosis with positive exponential growth. Either of these stable states may be reached under (some sets of) constant environmental conditions, depending on the initial values of symbiont and host biomass: if symbiont biomass is initially very low (i.e., a "bleached" coral), the system cannot escape the carbon-limited state and cannot grow. However, if symbiont biomass is initially high (i.e., a healthy coral), then the system maintains positive growth in a nitrogen-limited state. For practical purposes, we consider only positive growth steady states in the "steady state behavior" section below. In the subsequent "dynamic behavior" section, we explore more thoroughly the conditions that cause the system to switch between alternate stable states, and interpret these dynamics in the context of coral bleaching.

## Steady state behavior

@@

To analyze qualitative behavior, we ran the model to steady state across gradients of external irradiance and nutrients (Fig. 2), which revealed patterns consistent with observed phenomena in corals. Predicted growth rates are low at low light and DIN (~0.01 d^-1^), and begin increasing as both of these factors increase (Fig. 2A). Low light limits photosynthetic rates, resulting in less fixed carbon shared with the host and an associated increase in the symbiont to host biomass ratio (Fig. 2B). In agreement with this trend are many observations of negative correlation between irradiance and symbiont density [@Stimson:1997p3837; @Brown:1999p3534; @Fitt:2000p3792; @Titlyanov:2001p3631]. As higher light alleviates light-limitation of photosynthesis, growth becomes less carbon-limited.

@@

The maximum predicted growth rates of ~0.1 d^-1^, occurring between ~10-20 mol photons m^-2^ s^-1^ light and ~4 µM DIN (Fig. 2A), are comparable to the rate of `r Tanaka_gr` d^-1^ measured by @Tanaka:2007uv under similar N-enriched conditions. Predicted growth rates under conditions more typical of reef environments (<0.5 µM DIN) are ~0.01-0.03 d^-1^. Observed specific growth rates, which are typically based on skeletal mass, fall near or below 0.01 d^-1^ [@Osinga:2011kq; @Osinga:2012is] (though values as high as 0.025 d^-1^ have been reported [@Schutter:2010p7758]). That predicted growth rates are slightly higher than most observations is not surprising, as the model does not account for external ecological factors that may reduce growth, such as competition, predation, and bioerosion. Furthermore, the model does not explicitly account for skeletal formation, which may not always be well correlated with biomass growth, potentially due to differential substrate-limitation [@Anthony:2002tc]. Nevertheless, when rates of increase in coral tissue biomass have been directly measured over short periods, values are consistent with model predictions (e.g., `r Tanaka_gr` d^-1^ in @Tanaka:2007uv), and similar rates (`r Armoza_gr` d^-1^) have been measured in *Aiptasia diaphana*, a non-calcifying symbiotic anemone [@ArmozaZvuloni:2014ju].

@@

While bleaching in response to high light alone has been observed experimentally [@Schutter:2011ie; @Downs:2013kc], mass coral bleaching events occur in response to high temperature [@HoeghGuldberg:1999p4222]; thus, it is important to justify our consideration of light as the primary stressor in the model. In reality, light and temperature interact synergistically [@Coles:1978p1124; @Jones:1998p3572], and in fact, any stressor that disrupts the quenching of light energy may lead to bleaching [@Wooldridge:2010p7809; @Baker:2015kp]. This is because the proximate cause of coral bleaching is excess excitation energy, but the upstream events that lead to this situation may be diverse: indeed, elevated temperature may inhibit Rubisco functioning [@Jones:1998p3572] and the repair of the D1 protein in photosystem II [@Warner:1999p4239], which reduces the capacity of photochemical quenching and leads to an excess of light energy. In this way, elevated temperature serves to reduce the threshold above which light causes stress to the system [@HoeghGuldberg:1999p4222]; importantly, light is still the proximate stressor. Therefore, we omitted temperature from the model to maintain a desired level of simplicity, while still allowing photophysiological stress and bleaching to be simulated with biological realism in response to light. Thus, while the model is currently accurate from a conceptual and mechanistic standpoint, its value as a predictive tool under scenarios of changing temperature will require further development.

@@

## Sensitivity analysis

Each default parameter value used in the model (Table 2) is based on carefully selected relevant literature. The derivations of each of these default parameter values are described in the Supplementary Information. Nevertheless, it is important to evaluate the sensitivity of the model to changes in these parameter values. Moreover, these sensitivity plots serve to demonstrate the behavior of the dynamical system. We measured fractional change in steady state values in response to fractional changes in parameter values, relative to their default values, under environmental conditions typical of coral reefs.

@@

Overall, relative changes in the steady state of the system were less than the equivalent relative change in parameter value, indicating that the system maintains internal stability. However, changes in certain parameter values have greater impacts: increasing $j_{Nm}$ or decreasing $K_N$ both dramatically increase host growth, demonstrating the strong nitrogen-limitation that characterizes these symbioses (REF). The parameterr $\bar{a}^*$ has a strong impact on S:H biomass ratios, because this parameter determines the amount of light absorbed by symbionts, thus low values of this parameter increase light-limitation of the system. Altering the maximum growth and turnover rates of symbiont and host biomass also have expected impacts on steady states. Parameters having to do with photooxidative stress and bleaching have little impact under these environmental conditions, but of course have larger impacts under higher light (e.g., Fig. S6).

## Dynamic behavior
Therefore, we are highly interested in the conditions under which the system switches from one state to another, representing the conditions under which corals "bleach" due to progressive, reinforcing carbon-limitation. Interestingly, this state can be reached in multiple ways, by the acceleration of photoinhibition and photodamage, or by high external DIN, which causes the symbiont to share less carbon, subsequently causing its own CO2-limitation, and progressive C-limitation of the system.
## varying light: consequences of excess light and light-limitation
## varying DIN: role of N-limitation and consequences of elevated N
## elevated nutrients cause system to become C-limited (photo: CO2 limited; host: fixed-C limited)
## C-limitation increases susceptibility to bleaching
Importantly, these negative impacts of high irradiance only arise when light energy exceeds the capacity of photochemistry and NPQ, which in turn depends on CO~2~ availability. While previous models framed photophysiological stress as a fixed response to absolute irradiance levels [@Eynaud:2011tv], the current implementation considers the dynamic balance of multiple energy sinks in the causation of stress, which is more consistent with current understanding of symbiosis dysfunction [@Wooldridge:2013tj], and establishes an important role of the host in providing CO~2~ for symbiont photosynthesis [@Tansik:2015cm; @Hopkinson:2015ev].
## consequences of a less efficient but more tolerant symbiont
## Rate of symbiont repopulation (Berner et al. 1993)
## Bicarbonate addition promotes coral growth (Marubini and Thake 1999)
## Seasonal change in sym. density is larger when nutrients are elevated (Stimson 1997). Such a phenomenon was predicted by Fitt et al. 2000.
## Death rule: negative growth for a certain amount of time should be defensible.

>>>>

3. _Provide well-documented, open-access code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide open access to the model in the form of detailed, commented code written in the R language [@RALanguageandEn:2014wf]. With an accessible and modular framework, we envision this model as a resource for futher development by the scientific community to include additional complexity and problem-specific components. To widen the audience for this work, we chose the R language because it is freely available and in common use by biologists and ecologists.

@@

# Model description
In this dynamic bioenergetic model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1, and each model flux and parameter are defined in Tables 1 and 2, respectively. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Host biomass ($H$), symbiont biomass ($S$), and prey biomass ($X$) have fixed, but different, molar N:C ratios (Table 2). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the parallel complementary formulation of @Kooijman:2010vd to specify these fluxes. The two state variables are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area, and surface area is proportional to volume (i.e., corals are "V1-morphs" in DEB terminology [@Kooijman:2010vd]), biomass increases exponentially during growth (indeed, corals grow exponentially [@Bak:1976bv]).

Environmental stress is implemented in the form of photo-oxidative stress, which is thought to be the primary trigger of coral bleaching [@Lesser:1997p3587; @Weis:2008p944; @Wooldridge:2009p7807]. To simulate these events, we model the absorption and quenching of light energy by photochemistry and non-photochemical quenching, and the phenomenological responses that occur when these capacities are overwhelmed (i.e., photoinhibition, photodamage, and symbiont loss). While bleaching in response to high light alone has been observed experimentally [@Schutter:2011ie; @Downs:2013kc], mass coral bleaching events occur in response to high temperature [@HoeghGuldberg:1999p4222]; thus, it is important to justify our consideration of light as the primary stressor in the model. In reality, light and temperature interact synergistically [@Coles:1978p1124; @Jones:1998p3572], and in fact, any stressor that disrupts the quenching of light energy may lead to bleaching [@Wooldridge:2010p7809; @Baker:2015kp]. This is because the proximate cause of photo-oxidative stress is excess excitation energy, but the upstream events that lead to this situation may be diverse: indeed, elevated temperature may inhibit Rubisco functioning [@Jones:1998p3572] and the repair of the D1 protein in photosystem II [@Warner:1999p4239], which reduces the capacity of photochemical quenching and leads to an excess of light energy. In this way, elevated temperature serves to reduce the threshold above which light stresses the system [@HoeghGuldberg:1999p4222]; importantly, light is still the proximate stressor. Therefore, we omitted temperature from the model to maintain a desired level of simplicity, while still allowing photophysiological stress and bleaching to be simulated with biological realism in response to light.

Below we describe the formulation of each flux in the model.

@@

| Symbol  | Description                            | Units               | Eq. no. |
| ------- | -------------------------------------- | ------------------- | ------- |
| $j_X$     | Prey uptake rate                       | molX CmolH^-1^ d^-1^ | 3 |
| $j_N$     | Nitrogen uptake rate                   | molN CmolH^-1^ d^-1^ | 4 |
| $j_{HG}$  | Host biomass formation rate            | CmolH CmolH^-1^ d^-1^| 5 |
| $r_{NH}$  | Recycled nitrogen from host turnover   | molN CmolH^-1^ d^-1^ | 6 |
| $\rho_N$  | Nitrogen shared with the symbiont      | molN CmolH^-1^ d^-1^ | 7 |
| $j_{eC}$  | Excess carbon used to activate host CCMs| molC CmolH^-1^ d^-1^ |  |
| $j_L$     | Light absorption rate                  | molph CmolS^-1^ d^-1^| 9 |
| $j_{CO_2}$| CO~2~ input to photosynthesis          | molCO~2~ CmolH^-1^ d^-1^| 10 |
| $r_{CH}$  | Recycled CO~2~ from host turnover      | molCO~2~ CmolH^-1^ d^-1^| 11 |
| $r_{CS}$  | Recycled CO~2~ from symbiont turnover  | molCO~2~ CmolS^-1^ d^-1^| 12 |
| $j_{CP}$  | Photosynthesis rate                    | molC CmolS^-1^ d^-1^ | 13 |
| $j_{eL}$  | Light energy in excess of photochemistry| molph CmolS^-1^ d^-1^| 14 |
| $j_{NPQ}$ | Total capacity of NPQ                  | molph CmolS^-1^ d^-1^ | 15 |
| $c_{ROS}$ | ROS production proportional to baseline | -- | 16 |
| $r_{NS}$  | Recycled nitrogen from symbiont turnover| molN CmolS^-1^ d^-1^| 17 |
| $j_{SG}$  | Symbiont biomass formation rate        | CmolS CmolS^-1^ d^-1^| 18 |
| $\rho_C$  | Fixed carbon shared with host          | molC CmolS^-1^ d^-1^ | 19 |
| $j_{ST}$  | Symbiont turnover rate                 | CmolS CmolS^-1^ d^-1^| 20 |

@@

| Symbol  | Description                            | Units               | Value   |
| ------- | -------------------------------------- | ------------------- | ------- |
| $n_{NH}$ | N:C molar ratio in host biomass | -- | 0.19 |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | -- | 0.2 |
| $n_{NX}$ | N:C molar ratio in prey biomass | -- | 0.13 |
| $j_{HT}^0$ | Specific turnover rate of host biomass | CmolH CmolH^-1^ d^-1^ | 0.03 |
| $j_{ST}^0$ | Specific turnover rate of symbiont biomass | CmolS CmolS^-1^ d^-1^ | 0.03 |
| $\sigma_{NH}$ | Proportion host nitrogen turnover recycled | -- | 0.9 |
| $\sigma_{CH}$ | Proportion host carbon turnover recycled | -- | 0.9 |
| $\sigma_{NS}$ | Proportion symbiont nitrogen turnover recycled | -- | 0.9 |
| $\sigma_{CS}$ | Proportion symbiont carbon turnover recycled | -- | 0.9 |
| $j_{Xm}$ | Maximum specific feeding rate of host | molX CmolH^-1^ d^-1^ | 0.1292 |
| $K_X$ | Half-saturation constant for prey uptake by host | molX L^-1^ | 20e-6 |
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | molN CmolH^-1^ d^-1^ | 0.048 |
| $K_N$ | Half-saturation constant for DIN uptake by host | molN L^-1^ | 0.46e-6 |
| $k_{CO_2}$ | Efficacy of CO~2~ delivery by host CCMs | molCO~2~ molC^-1^ | 10 |
| $j_{HGm}$ | Maximum specific growth rate of host | CmolH CmolH^-1^ d^-1^ | 1 |
| $y_{CL}$ | Quantum yield of photosynthesis | molC mol ph^-1^ | 0.1 |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | m^2^ CmolS^-1^ | 1.34 |
| $k_{NPQ}$ | NPQ capacity of symbiont | mol ph CmolS^-1^ d^-1^ | 112 |
| $k_{ROS}$ | Excess photon energy that doubles ROS production, relative to baseline levels | mol ph CmolS^-1^ d^-1^ | 40-80 |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | molC CmolS^-1^ d^-1^ | 2.8 |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | CmolS CmolS^-1^ d^-1^ | 0.25 |
| $b$ | Scaling parameter for strength of bleaching response | -- | 5 |

@@

We then specify the input of CO~2~ to photosynthesis as $j_{CO_2}$, a flux mediated by the host that encompasses potentially diverse carbon-concentrating mechanisms (CCMs), including active transport of bicarbonate, carbonic anhydrase-catalyzed conversion of bicarbonate to CO~2~ to promote diffusion toward the symbiont [@Tansik:2015cm], and acidification of the symbiosome to increase localized CO~2~ concentrations around the symbiont [@Barott:2014gx]. Since these active CCMs require energetic input by the host, the magnitude of CO~2~ delivery is proportional to the amount of fixed carbon available to the host after satisfying its growth demands, $j_{eC}$, assuming that some of this carbon is respired to energize these CCMs. This formulation means that the symbiont indirectly ensures its own CO~2~ supply by providing fixed carbon (=energy) to the host [@Wooldridge:2013tj], which establishes a positive feedback in the model that leads to the existence of multiple stable states, which we discuss below in the context of coral bleaching. The parameter $k_{CO_2}$ scales the efficacy of host CCMs, which enables the comparison of different rates of CO~2~ delivery that may characterize different coral species [@Wooldridge:2014di]. The input of CO~2~ to the photosynthesis SU is therefore specified as:

@@

# Steady state behavior
In a constant environment, the system ultimately reaches a steady state of exponential growth or decline. However, under some conditions, either of these outcomes may occur depending on initial values of symbiont and host biomass, indicating the presence of alternate stable states. The mechanism that produces these alternate stable states is the positive feedback between carbon-limitation of the host and CO~2~-limitation of photosynthesis: if symbiont biomass is initially very low (i.e., a "bleached" coral), the system cannot escape positively reinforcing carbon-limitation and cannot grow. However, if symbiont biomass is initially high (i.e., a "healthy" coral), then the system maintains positive growth in a nitrogen-limited state. For practical purposes, this section of the manuscript considers only positive growth steady states under constant environments; the subsequent "dynamic behavior" section explores how changing environments may cause the system to switch between alternate stable states, which we interpret in the context of coral bleaching.

To analyze qualitative patterns in steady state behavior, we ran the model to steady state across gradients of external irradiance and nutrients (Fig. 2), which revealed patterns consistent with observed phenomena in corals. Predicted growth rates are low at low light and DIN (~0.01 d^-1^), and begin increasing as both of these factors increase (Fig. 2A). Low light limits photosynthetic rates, resulting in less fixed carbon shared with the host and an associated increase in the symbiont to host biomass ratio (Fig. 2B). In agreement with this trend are many observations of negative correlation between irradiance and symbiont density [@Stimson:1997p3837; @Brown:1999p3534; @Fitt:2000p3792; @Titlyanov:2001p3631]. As higher light alleviates light-limitation of photosynthesis, growth becomes less carbon-limited.

@@

@@

The maximum predicted growth rates of ~0.1 d^-1^, occurring between ~10-20 mol photons m^-2^ s^-1^ light and ~4 µM DIN (Fig. 2A), are comparable to the rate of `r Tanaka_gr` d^-1^ measured by @Tanaka:2007uv under similar N-enriched conditions. Predicted growth rates under conditions more typical of reef environments (<0.5 µM DIN) are ~0.01-0.03 d^-1^. Observed specific growth rates, which are typically based on skeletal mass, fall near or below 0.01 d^-1^ [@Osinga:2011kq; @Osinga:2012is], though values as high as 0.025 d^-1^ have been reported [@Schutter:2010p7758]. It is not surprising that predicted growth rates are slightly higher than most observations, as the model does not account for external ecological factors that may reduce growth, such as competition, predation, and bioerosion. Furthermore, the model does not explicitly account for skeletal formation, which may not always correlate with biomass growth, potentially due to differential substrate-limitation [@Anthony:2002tc]. Nevertheless, when growth of coral tissue biomass has been directly measured over short periods, values are consistent with model predictions (e.g., `r Tanaka_gr` d^-1^ in @Tanaka:2007uv), and similar rates (`r Armoza_gr` d^-1^) have been measured in *Aiptasia diaphana*, a non-calcifying symbiotic anemone [@ArmozaZvuloni:2014ju].

@@

@@

# Sensitivity analysis

The values used for each parameter in the model (Table 2) are derived from relevant literature; these derivations are described in the Supplementary Information. Here we evaluate the sensitivity of the model to changes in these parameter values, which also serves to demonstrate the behavior of the dynamical system. We measured fractional change in steady state values in response to fractional changes in parameter values, relative to their default values, under environmental conditions typical of coral reefs.

@@

Overall, relative changes in the steady state of the system are less than the equivalent relative change in parameter value, indicating that the system maintains internal stability. However, changes in certain parameter values have more significant impacts: increasing $j_{Nm}$ or decreasing $K_N$ both dramatically increase host growth (Fig. 3), demonstrating the strong nitrogen-limitation that characterizes these symbioses (REF). The parameter $\bar{a}^*$ has a strong impact on S:H biomass ratios (Fig. 3) since this parameter determines the amount of light absorbed by symbionts, with lower values increasing light-limitation. Altering maximum growth and turnover rates of symbiont and host also have expected impacts on steady states (Fig. 3). Parameters relating to photooxidative stress and bleaching have little impact under these environmental conditions (Fig. 3), but of course have larger impacts under higher light (e.g., Fig. S6). Sensitivity analyses conducted under different combinations of external light and nutrients are presented in the Supplementary Information (Figs. S3-S7)

# Dynamic behavior
The dynamic behavior of the model demonstrates its power to integrate multiple environmental forcings simultaneously to produced integrated responses of the system. The potential applications of this are wide-ranging, and so the model may be highly useful for generating hypotheses about the way the system will respond to complex sets of environmental dynamics. Here we present several scenarios to highlight the power of the model and demonstrate that it can also reproduce more complex phenomena that have been observed in corals.

## Seasonal variability
Symbiont densities and coral growth rates are known to vary seasonally, representing an integrated response to changes in a suite of environmental factors. Light in particular is a strong driver of these trends [Stimson:1997p3837; @Brown:1999p3534; @Fagoonee:1999p4136; @Fitt:2000p3792], with high light associated with lower symbiont abundance and reduced tissue biomass. The role of light in driving seasonal changes in symbiont density was demonstrated nicely by @Stimson:1997p3837, who also found that experimental nutrient-enrichment amplified the light-driven seasonal oscillation. Using the levels of light and nutrients from this study as inputs, the model reproduces this observed interaction among environmental factors (Fig. 4), and also provides the mechanism: under high light in summer, photoinhibition reduces symbiont biomass and leads to carbon-limitation of host growth regardless of nutrient status; when light is reduced in winter, these responses become constrained instead by nitrogen, and therefore increase more when DIN is enriched (Fig. S8).

![Figure 4. Light-driven seasonal dynamics of symbiont abundance and coral growth. Light input (upper panel) was designed as a sinusoidal curve with a period of one year, with maximum and minimum values of 40 and 22 mol photons m^-2^ d^-1^, based on @Stimson:1997p3837. The dynamic behavior of symiont to host biomass ratio (middle panel) and the specific growth rate of host biomass (lower panel) show seasonal oscillations that are greater in magnitude under higher nutrients (1e-5 molN L^-1^; dashed lines) relative to lower nutrients (1e-7 molN L^-1^; solid lines), consistent with the findings of @Stimson:1997p3837. Prey density was set at 1e-5 CmolX L^-1^.](../img/Fig4.png)

The prediction of higher growth when light is reduced indicates that growth is not limited by low light in winter, but is actually reduced by excess light in summer, consistent with the experimental findings of @Schutter:2011ie. Seasonal summertime reductions in tissue biomass have also been well-documented in the field [@Fitt:2000p3792], along with reductions in net photosynthetic capacity [@MullerParker:1987cr]. Importantly, while light alone (at these levels) may drive seasonal dynamics in the ways discussed, temperature fluctuations may attenuate or even reverse the effect of light as cooler winters depress metabolism; thus, the relative magnitude of fluctuation in temperature, light, and other factors may produce wide variability in the direction and magnitude of seasonal changes in growth and symbiont abundance, depending on location and microhabitat. Nevertheless, the seasonal variability predicted here (Fig. 4) is consistent with experimental and field observations for corals, and demonstrates the power of this model to predict dynamic behavior that mechanistically integrate the impacts of multiple environmental drivers.

## Coral bleaching
Coral bleaching is the stress-induced loss of symbiotic algae from coral tissues, which can occur in response to a variety of environmental stressors. In most cases, coral bleaching is thought to begin with photo-oxidative stress in symbiont photosynthesis [@Lesser:1997p3587], which triggers a cascade of events leading to symbiont expulsion [@Weis:2008p944]. As symbionts are expelled, the host receives less fixed carbon, which may then compromise its ability to deliver CO~2~ to photosynthesis [@Wooldridge:2013tj]. Increasing CO~2~-limitation for remaining symbionts, along with an amplified internal light environment due to reduced self-shading [@Enriquez:2005p142], may further exacerbate photo-oxidative stress and accelerate symbiont expulsion, driving the coral into a bleached state.

While these positive feedbacks have been discussed previously in the literature, this is the first attempt to implement and explore their properties within a dynamical model. Interestingly, these feedbacks lead to alternate stable states. The 'healthy' state is characterized by nitrogen-limitation of both symbiont and host: under these conditions, the symbiont translocates sufficient carbon to support host growth and CCMs, which ensures that photosynthesis does not become CO~2~-limited. However, if carbon translocation is disrupted (and light is sufficiently high), then the system is driven into the 'bleached' state by photo-oxidative stress and positively reinforcing carbon- and CO~2~-limitation. In this context, coral bleaching can be understood as a transition from one stable state to another, and bleaching thresholds are sets of environmental conditions that push a healthy-state coral onto a trajectory leading to a bleached stable state.

We are highly interested in the conditions under which the system switches from a healthy to a bleached state, and can use this model as a tool to explore this dynamic behavior. Most straightforwardly, this switch occurs when irradiance exceeds the capacity of photochemistry and NPQ, causing the production of ROS that leads to photoinhibition, photodamage, and symbiont expulsion, thereby engaging the positive feedbacks that lead to a bleached state (Fig. 5). However, the positive feedbacks involved in bleaching are not engaged only in response to high external irradiance; in fact, they depend on the relative balance of light energy absorption and quenching, which in turn depends on the availability of CO~2~ for photosynthesis. While previous models framed photo-oxidative stress as a fixed response to absolute external irradiance [@Eynaud:2011tv], our implementation considers the dynamic balance of multiple energy sinks in the causation of stress, which is more consistent with current understanding of symbiosis dysfunction [@Wooldridge:2013tj], and establishes a critical role of host CCMs in providing CO~2~ for photosynthesis [@Tansik:2015cm; @Hopkinson:2015ev].

![Figure 5. Bleaching.](../img/Fig5.png)

The importance of host CCM activity establishes significant interactive roles for other factors in influencing coral bleaching responses. For example, simulations of high light stress (Fig. 6) demonstrate that bleaching can be attenuated by heterotrophic feeding, a phenomenon which has been observed experimentally [@Borell:2008p108]. The mechanism underlying this prediction is that feeding by the host increases host CCM activity, which delays the onset of CO~2~-limitation of photosynthesis and reduces bleaching severity. On the other hand, elevated nutrients exacerbate bleaching (Fig. 6), since higher symbiont densities are more susceptible to CO~2~-limitation [@Wooldridge:2009p7807]. Several experimental [@Cunning:2013gp; @VegaThurber:2014hk] and correlational studies [@Wooldridge:2009kb] are consistent with this mechanistic link between high nutrients and bleaching.

![Figure 6. Bleaching and interactive factors.](../img/Fig6.png)

Since bleaching represents an alternative stable state, it follows that a bleached coral cannot recover unless the external condition that induced bleaching is alleviated. In natural settings, this typically occurs by seasonal declines in temperature and light. However, due to hysteresis, a coral cannot recover along the same trajectory it followed during bleaching; indeed, the stressor must be alleviated well *below* the threshold that initially caused bleaching in order for the system to recover (Fig. 7). This is because under the same external conditions, a bleached coral (relative to a healthy one) is characterized by greater light amplification and weaker CCM activity, which serve to maintain the bleached state. The magnitude of hysteresis will depend on external nutrients and prey availability, as well as model parameters, yet the fundamental concept demonstrates a core behavior of the system.

![Figure 7. Hysteresis.](../img/Fig7.png)

The hysteresis demonstrated in Fig. 7 reveals another interesting behavior of the system: during recovery from bleaching, the S:H biomass ratio reaches a higher value that 'overshoots' the level maintained in the healthy stable state. In fact, such a phenomenon has been observed in both experimental [@Cunning:2015ja] and field studies [@Kemp:2014ug] documenting recovery from bleaching, and has been interpreted as a potential 'disequilibrium in host-symbiont regulation' [@Kemp:2014ug]. The model explains this 'overshoot' as an approach to a bifurcation point in the dynamical system: during recovery from the 'bleached' state, host growth is more strongly carbon-limited than symbiont growth (since the symbiont has priority access to carbon), so S:H biomass increases. S:H biomass will continue to increase beyond the alternative steady state levels as long as the symbiont remains carbon-limited; not until nitrogen becomes more strongly limiting to symbiont growth (hence forcing the symbiont to share fixed carbon with the host) will the system 'flip' back to a state of high carbon fixation and overall nitrogen-limitation that characterize a 'healthy' symbiosis. Of course, the magnitude of this 'overshoot' will depend on external nutrient concentrations and the rate of change in the stressor during recovery, though simulations predict similar magnitudes to what was observed by Kemp et al. (supp fig?).
